A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-26 latex/main.tex
Update

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
Boxes identify specific samples, colored by their Pango X classification;
smaller squares list extra samples descending from a node, hidden for brevity,
with all hidden samples taking the same Pango designation as their parent.
For the two recombination nodes, a genome bar above shows
the location of the inferred breakpoint interval relative to the Spike gene.
For simplicity, ancestral lineages above the recombination parents or the MRCA are omitted.
(A) Subgraph around the recombination node associated with XA.
Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages: we use the label Xx as a shorthand for this related group of recombinants; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

[TODO narrative is muddled here: the point of the para is to say that 
pango X lineage designation is manual and subjective.]
% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis. 
The first recombinant lineage XA was designated in 2021 based on the 
analysis of Jackson et al\.~\cite{Jackson2021}, which established 
that recombination with onward transmission had occured [weak FIXME].
As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146
Although the process of incorporating recombinant lineages in the Pango 
nomenclature system is well defined,
the identification of such events is much more subjective, requiring
the synthesis of multiple lines of evidence and often the 
manual inspection of phylogenies and sequence alignments
by volunteers.

By automatically detecting and incorporating recombination events
into the ARG, sc2ts provides us with a systematic means of 
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarises
these results for the 44 Pango X lineages present in
the ARG (XAC and XAD have been omitted, and XM 
partially represented, corresponding to 23 samples in total; see STAR Methods).
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent (see next 
section). 13 of the Type I events are monophyletic for the focal
pango (i.e., the descendants are entirely from that lineage); 
Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
Document~\ref{sec:jackson_recombs} for further analysis 
of the Jackson et al.\ recombinants).

@@

\subsubsection*{Filtering time travellers}
Time-travelling samples---those in which the recorded collection date differs
substantially from their true collection date---pose significant problems
for sc2ts, and are present at a significant frequency. For example, of the 
43 Scorpio designations present in the Viridian dataset studied, 32 have 
samples present with sampling dates in 2020, including lineages that do not
arise until late 2022. Inserting these time-travellers into the ARG 
at the claimed sampling date causes major issues, and they must therefore be 
filtered out. We use the simple approach of filtering out samples that 
exceed a given ``HMM cost'', i.e., requiring too many mutations or
recombination breakpoints to plausibly represent the steady accumulation
of diversity expected with dense sampling. 
The HMM cost for the Viterbi solution for a sample under the LS model
is $k$ times the number of recombination events plus the number of 
mismatches, where $k$ is the recombination penalty (here, $k=4$).
After some experimentation, we set the recombination penalty to 7, 
disallowing samples that require 8 or more mutations (or, e.g., 1 recombination
and 4 mutations, or 2 recombinations) from immediate inclusion (see next
section) in the ARG. This approach has the benefit of filtering
samples enriched for errors as well as time-travellers.

@@

\subsection{Analysis of Pango lineage ``origination'' events}
\label{sec:supplement_pango_lineages}
To analyze how well the ARG reflects the phylogenetic structure implicit in the
Pango lineage naming system, we considered the Pango assignments generated by
Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
perfectly reflected the Pango lineage structure, each lineage would form a
clade descending from a single originating node with that label. We identified
putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there
is a tie).

@@

>>>>

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
Boxes identify specific samples, colored by their Pango X classification;
smaller squares list extra samples descending from a node, hidden for brevity,
with all hidden samples taking the same Pango designation as their parent.
For the two recombination nodes, a genome bar above shows
the location of the inferred breakpoint interval relative to the Spike gene.
For simplicity, ancestral lineages above the recombination parents or the MRCA are omitted.
(A) Subgraph around the recombination node associated with XA.
Two ``Group A'' recombinant samples reported by Jackson 
et al.\cite{Jackson2021} are shown.
(B) Subgraph around the recombination node associated with a 
group of ``nested'' Pango X lineages: we use the 
label Xx as a shorthand for this related group of recombinants; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

[TODO narrative is muddled here: the point of the para is to say that 
pango X lineage designation is manual and subjective.]
% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis. 
The first recombinant lineage XA was designated in 2021 based on the 
analysis of Jackson et al.\cite{Jackson2021}, which established 
that recombination with onward transmission had occured [weak FIXME].
As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146
Although the process of incorporating recombinant lineages in the Pango 
nomenclature system is well defined,
the identification of such events is much more subjective, requiring
the synthesis of multiple lines of evidence and often the 
manual inspection of phylogenies and sequence alignments
by volunteers.

By automatically detecting and incorporating recombination events
into the ARG, sc2ts provides us with a systematic means of 
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarises
these results for the 44 Pango X lineages present in
the ARG (XAC and XAD have been omitted, and XM 
partially represented, corresponding to 23 samples in total; 
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent (see next 
section). 13 of the Type I events are monophyletic for the focal
pango (i.e., the descendants are entirely from that lineage); 
Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
Document~\ref{sec:jackson_recombs} for further analysis 
of the Jackson et al.\ recombinants).

@@

\subsubsection*{Filtering time travellers}
Time-travelling samples---those in which the recorded collection date differs
substantially from their true collection date---pose significant problems
for sc2ts, and are present at a significant frequency. For example, of the 
43 Scorpio~\cite{Colquhoun2023scorpio}
designations present in the Viridian dataset studied, 32 have 
samples present with sampling dates in 2020, including lineages that do not
arise until late 2022. Inserting these time-travellers into the ARG 
at the claimed sampling date causes major issues, and they must therefore be 
filtered out. We use the simple approach of filtering out samples that 
exceed a given ``HMM cost'', i.e., requiring too many mutations or
recombination breakpoints to plausibly represent the steady accumulation
of diversity expected with dense sampling. 
The HMM cost for the Viterbi solution for a sample under the LS model
is $k$ times the number of recombination events plus the number of 
mismatches, where $k$ is the recombination penalty (here, $k=4$).
After some experimentation, we set the recombination penalty to 7, 
disallowing samples that require 8 or more mutations (or, e.g., 1 recombination
and 4 mutations, or 2 recombinations) from immediate inclusion (see next
section) in the ARG. This approach has the benefit of filtering
samples enriched for errors as well as time-travellers.

@@

\subsection{Pango lineage ``origination'' events}
\label{sec:supplement_pango_lineages}
To analyze how well the ARG reflects the phylogenetic structure implicit in the
Pango lineage naming system, we considered the Pango assignments generated by
Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
perfectly reflected the Pango lineage structure, each lineage would form a
clade descending from a single originating node with that label. We identified
putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there
is a tie).

@@

\subsection{Pango X lineage events}
\label{sec:supplement_x_pango_lineages}
[TODO DOCUMENT DETAILS of deriving events]
